Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-12 paper-1/manuscript.tex
Replaced V_{TA} with \theta (text)

When a neuron's membrane potential reached the firing threshold $V_\theta$, a spike was emitted, and the potential was clamped to $V_m = V_{reset} = -74$~mV for a fixed refractory period of 3~ms (excitatory neurons). Inhibitory neurons were modeled as fast-spiking cells with a refractory period of 2~ms and a constant firing threshold $V_\theta = \theta_0 = -54$~mV \citep{Mensi2012-au}. In contrast, excitatory neurons were modeled with threshold adaptation (TA) as in \cite{Teeter2018-iz}, following
\begin{align}
    V_\theta &= \theta_0 + V_{TA}(t) \nonumber \\
    \tau_{\theta} \frac{dV_{TA}}{dt} &= -V_{TA} + \hat{\theta} \delta(t - \hat{t}) \label{eqn-TA}
\end{align}
with increment $\hat{\theta} = 1$~mV and decay time constant $\tau_{\theta} = 1$~s \citep{Pozzorini2015-ei}.

@@

In order to disentangle the effects of STD and TA, we ran all simulations in four conditions: Without any short-term plasticity, with either TA only or STD only, and with both STD and TA (labeled STD+TA in the following, and corresponding to the full model as described above). To turn off STD, we replaced Equation \ref{eqn-xsyn} with a constant $x = 1$, but retained the scaling of inputs to $g_e$ with $U$ in Equation \ref{eqn-gsyn}. This corresponds to immediate recovery ($\tau_x = 0$) and was done to maintain the magnitude of the excitatory postsynaptic potential (EPSP) in the recovered state regardless of whether STD was turned on or off. To turn off TA, we replaced Equation \ref{eqn-TA} with a constant $V_{TA} = 0$, making the firing threshold completely unadaptive.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower $V_{TA}$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (\dev{} - \msc{}) in $V_{TA}$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Network-wide median contrast (\dev{} - \msc{}) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (two-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

We quantified the former with the strength of adaptation $V_{TA}$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of target \dev{} trials relative to \msc{}, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). On the other hand, once the bulk of activity had occurred, many neurons showed higher thresholds in \dev{} (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $V_{TA}$, leading to numerous horizontal breaks in the plot.

To assess the difference in synaptic currents in a way that allowed direct comparison to $V_{TA}$, we referred to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $V_{TA}$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We correlated the level of $V_{TA}$ at the start of target trials (i.e., at the left edge of the histograms in Figure \ref{fig4}b) with the average target trial response in terms of spikes fired across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median $V_{TA}$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the neuron-level correlation between the contrasts in $V_{TA}$ and target response was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $V_{TA}$ in \dev{} tended to respond more strongly to target stimuli.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $V_{TA}$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $V_{TA}$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $V_{TA}$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $V_{TA}$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $V_{TA}$ in \dev{} is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $V_{TA}$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages for \dev{} and \msc{} sequences, respectively; right column, contrast between \dev{} and \msc{} averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials (B in \dev{}, B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between the contrast (\dev{} - \msc{}) in average number of spikes fired in response to non-target stimulation and the contrast in $V_{TA}$ at the start of target trials, and associated linear regression.
\textbf{d} Network-wide median contrast (\dev{} - \msc{}) in the non-target response magnitude across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel (non-target response size contrast vs $\Delta V_{TA}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

First, we mapped the average value of $V_{TA}$ at the start of target trials to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the \msc{} sequence, $V_{TA}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the \dev{} sequence, neurons near the non-target stimulation site (B) were strongly adapted, whereas the majority of the network was less adapted, consistent with the data shown above.

@@

Correlating the non-target response contrast (\dev{} - \msc{}) with the contrast in $V_{TA}$ across excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in \dev{} almost perfectly predicted the resulting reduction in TA levels before target stimuli (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.99 and a consistent reduction in the median non-target response between \msc{} and \dev{} sequences (T = 525, p = 2.12e-16, median = -0.094; one-sided Wilcoxon, n = 120), clearly validating our notion that the reduced non-target trial response was the primary driver of lower TA, which we showed above to be responsible for the higher target response.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between the non-target response contrast (\dev{} - \msc{}) and the contrast in $V_{TA}$ at the start of B trials, and associated linear regression across all excitatory neurons. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
\textbf{c} Median non-target response contrast (\dev{} - \msc{}) among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labeled "late", across networks and stimuli.
\textbf{c} Median contrast (\dev{} - \msc{}) in $V_{TA}$ among the early and late neurons, respectively, across networks and stimuli.
\textbf{d} Relationship across networks and stimuli between the median contrast in $V_{TA}$ among the early portion and the median contrast in the response to stimulation in B, and associated linear regression

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we found, for each neuron, the minimum response time (i.e., time to first spike) across B trials in \dev{}, \msc{} and \std{} sequences, and ranked neurons by this response time. The ranking in the sample network is shown in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast (\dev{} - \msc{}) to the contrast in $V_{TA}$ at the start of B trials. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in Figure \ref{fig5}c, where we correlated the same response contrast to $\Delta V_{TA}$ before A. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the \dev{} sequence, while the response of the later portion of the network decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that the non-target response and $V_{TA}$ contrasts were robustly positively correlated (median Pearson's $\rho$ 0.992, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast (Figure \ref{fig6}c) and the median contrast in $V_{TA}$ at the start of B trials (Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in \dev{} than \msc{} (T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). The contrast in $V_{TA}$ followed the same pattern, with the early portion exhibiting stronger TA in \dev{} (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).

Finally, to show that early $V_{TA}$ was causal in reducing the response to B, we correlated the median contrast in $V_{TA}$ among the early portion of neurons with the network-wide median contrast of the response to stimulation in B, as shown in Figure \ref{fig6}e. Across networks and stimuli, we found a strong negative correlation ($\rho$ = -0.587, p = 9.56e-13, Wald test, n = 120), indicating that greater TA in the neurons responding early was indeed predictive of a lowered response to B in the overall network.

To sum up, we showed in Figure \ref{fig4} that greater $V_{TA}$ in the portion of the network responding earliest to a stimulus plays an important role in reducing the overall response. Here, we showed that the neurons responding earliest to B exhibit higher $V_{TA}$ in the \dev{} sequence as a consequence of an increased response, and that this increase in TA was directly responsible for reducing the response size and TA levels of the network as a whole.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model maintains or slighly enhances the effects of TA.
\textbf{a} Median contrast (\dev{} - \msc{}) in the non-target response (left) and in $V_{TA}$ at the start of B trials (right) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to Figure \ref{fig6}c-d.
\textbf{b} Median contrast (\dev{} - \msc{}) in the non-target response magnitude across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (\dev{} - \msc{}) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{c} Non-target response contrast, plotted against its Pearson correlation coefficient with the corresponding contrast in $V_{TA}$ at the start of target trials, in the full model. Compare to Figure \ref{fig5}d.
\textbf{d} Median contrast (\dev{} - \msc{}) in $V_{TA}$ at the start of target trials across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that $V_{TA}$ tends to be reduced more in the full model.
\textbf{e} $\Delta V_{TA}$ at the start of target trials, plotted against its Pearson correlation coefficient with the corresponding contrast in the target response magnitude, in the full model. Compare to Figure \ref{fig4}e.
\textbf{f} Median contrast (\dev{} - \msc{}) in the target response magnitude across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model

To understand how STD affects deviance detection and causes this difference in response reduction, we first examined the effects of TA in the full model, following the analysis on the ablated model laid out in the previous section. Starting at the cause of the reduced response in non-target trials, we show in Figure \ref{fig7}a the full model equivalent of Figure \ref{fig6}c and d. The contrasts (\dev{} - \msc{}) in the non-target response magnitude and in the resulting $V_{TA}$ levels at the start of B trials both retained a clear separation between the portion of neurons responding early, which showed increased responses (T = 5.44e+03, p = 1.02e-06; one-sided Wilcoxon, n = 120) and increased TA levels (T = 3.41e+03, p = 9.51e-08), and the neurons responding later, whose response (T = 282, p = 9.19e-19) and TA levels (T = 72, p = 2.49e-17) decreased in the \dev{} sequence. Likewise, the correlation across excitatory neurons between these two contrasts remained positive (T = 7.26e+03, p = 9.86e-22) and very strong (median Pearson's $\rho$ = 0.992). This indicates that local suppression to B mediated by TA remained a key driver of the reduced response to non-target trials.

>>>>

When a neuron's membrane potential reached the firing threshold $V_\theta$, a spike was emitted, and the potential was clamped to $V_m = V_{reset} = -74$~mV for a fixed refractory period of 3~ms (excitatory neurons). Inhibitory neurons were modeled as fast-spiking cells with a refractory period of 2~ms and a constant firing threshold $V_\theta = \theta_0 = -54$~mV \citep{Mensi2012-au}. In contrast, excitatory neurons were modeled with threshold adaptation (TA) as in \cite{Teeter2018-iz}, following
\begin{align}
    V_\theta &= \theta_0 + \theta(t) \nonumber \\
    \tau_{\theta} \frac{d\theta}{dt} &= -\theta + \hat{\theta} \delta(t - \hat{t}) \label{eqn-TA}
\end{align}
with increment $\hat{\theta} = 1$~mV and decay time constant $\tau_{\theta} = 1$~s \citep{Pozzorini2015-ei}.

@@

In order to disentangle the effects of STD and TA, we ran all simulations in four conditions: Without any short-term plasticity, with either TA only or STD only, and with both STD and TA (labeled STD+TA in the following, and corresponding to the full model as described above). To turn off STD, we replaced Equation \ref{eqn-xsyn} with a constant $x = 1$, but retained the scaling of inputs to $g_e$ with $U$ in Equation \ref{eqn-gsyn}. This corresponds to immediate recovery ($\tau_x = 0$) and was done to maintain the magnitude of the excitatory postsynaptic potential (EPSP) in the recovered state regardless of whether STD was turned on or off. To turn off TA, we replaced Equation \ref{eqn-TA} with a constant $\theta = 0$, making the firing threshold completely unadaptive.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower $\theta$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $\theta$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (\dev{} - \msc{}) in $\theta$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Network-wide median contrast (\dev{} - \msc{}) in $\theta$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \theta$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (two-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $\theta$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $\theta$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $\theta$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

We quantified the former with the strength of adaptation $\theta$ (Equation \ref{eqn-TA}), i.e., the amount by which the firing threshold was raised due to TA. As shown in Figure \ref{fig4}b, there was a clear and widespread lowering of thresholds at the start of target \dev{} trials relative to \msc{}, visible both in the raw data (left and middle panels) and in the contrast (right panel, early blue portions). On the other hand, once the bulk of activity had occurred, many neurons showed higher thresholds in \dev{} (right panel, later orange portions), since they spiked more often. Finally, note that inhibitory neurons, which were modeled without TA, naturally showed no difference in $\theta$, leading to numerous horizontal breaks in the plot.

To assess the difference in synaptic currents in a way that allowed direct comparison to $\theta$, we referred to the membrane voltage $V_m$, which is largely independent from and complementary to the firing threshold, and driven mostly by presynaptic inputs, resets after spikes notwithstanding. As the immediate cause of neuronal spiking, the overall pattern of $V_m$ (Figure \ref{fig4}c) largely reflected that seen in spiking activity. The contrast (right panel) shows not only a notable relative increase in membrane voltages in \dev{} around the typical response time (note the orange band running along the response front), but also an increased hyperpolarization starting immediately after due to post-spike resets.

Based on these measures, we then asked why there were more spikes in \dev{} trials. We first focused on the contribution of $\theta$ alone in an effort to confirm that thresholds are lower in \dev{}, and that this reduction in thresholds causes increased firing. We correlated the contrast (dev - msc) of $\theta$ at the start of target trials (i.e., at the left edge of the contrast histogram in Figure \ref{fig4}b) with the target trial response size contrast, across excitatory neurons. In the sample network (Figure \ref{fig4}d), we found that thresholds were clearly lower in \dev{} than \msc{} (T = 2.69e+04, p = 1.58e-92, median = -0.166 mV; one-sided Wilcoxon, n = 800 excitatory neurons), and that this decrease was accompanied by an increase in the target trial response at the level of individual neurons (Pearson's $\rho$ = -0.424, p = 1.6e-36; one-sided Wald test, n = 800). As shown in Figure \ref{fig4}e, this relationship held across networks and stimuli: The median $\theta$ among excitatory neurons was lower in \dev{} than \msc{} (T = 514, p = 1.67e-16, grand median = -0.094 mV; one-sided Wilcoxon, n = 120), indicating that networks were less adapted during oddball sequences. In addition, the neuron-level correlation between the contrasts in $\theta$ and target response was almost universally negative (T = 40, p = 2.68e-21, median = -0.357). This indicates that neurons with a greater decrease in $\theta$ in \dev{} tended to respond more strongly to target stimuli.

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $\theta$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $\theta$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $\theta$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $\theta$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig5.pdf}
    \caption{}
    \label{fig5}
\end{figure*}
\textbf{Fig. 5} Lower $\theta$ in \dev{} is a consequence of lower \dev{} non-target response magnitude.
\textbf{a} $\theta$ at the start of A trials (mean over trials) in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages for \dev{} and \msc{} sequences, respectively; right column, contrast between \dev{} and \msc{} averages. A and B stimulus locations for this and following figures are highlighted in the left plot; see Figure \ref{fig1} for the locations of the remaining stimuli used in the \msc{} sequence.
\textbf{b} Average response (in spikes per trial) to non-target trials (B in \dev{}, B, C, D, and E in \msc{}). Columns are arranged as in panel a.
\textbf{c} Relationship across excitatory neurons of the sample network between the contrast (\dev{} - \msc{}) in average number of spikes fired in response to non-target stimulation and the contrast in $\theta$ at the start of target trials, and associated linear regression.
\textbf{d} Network-wide median contrast (\dev{} - \msc{}) in the non-target response magnitude across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel (non-target response size contrast vs $\Delta \theta$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations

First, we mapped the average value of $\theta$ at the start of target trials to the spatial location of the neurons, as shown in Figure \ref{fig5}a. In the \msc{} sequence, $\theta$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the \dev{} sequence, neurons near the non-target stimulation site (B) were strongly adapted, whereas the majority of the network was less adapted, consistent with the data shown above.

@@

Correlating the non-target response contrast (\dev{} - \msc{}) with the contrast in $\theta$ across excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in \dev{} almost perfectly predicted the resulting reduction in TA levels before target stimuli (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.99 and a consistent reduction in the median non-target response between \msc{} and \dev{} sequences (T = 525, p = 2.12e-16, median = -0.094; one-sided Wilcoxon, n = 120), clearly validating our notion that the reduced non-target trial response was the primary driver of lower TA, which we showed above to be responsible for the higher target response.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between the non-target response contrast (\dev{} - \msc{}) and the contrast in $\theta$ at the start of B trials, and associated linear regression across all excitatory neurons. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
\textbf{c} Median non-target response contrast (\dev{} - \msc{}) among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labeled "late", across networks and stimuli.
\textbf{c} Median contrast (\dev{} - \msc{}) in $\theta$ among the early and late neurons, respectively, across networks and stimuli.
\textbf{d} Relationship across networks and stimuli between the median contrast in $\theta$ among the early portion and the median contrast in the response to stimulation in B, and associated linear regression

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we found, for each neuron, the minimum response time (i.e., time to first spike) across B trials in \dev{}, \msc{} and \std{} sequences, and ranked neurons by this response time. The ranking in the sample network is shown in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast (\dev{} - \msc{}) to the contrast in $\theta$ at the start of B trials. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in Figure \ref{fig5}c, where we correlated the same response contrast to $\Delta \theta$ before A. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the \dev{} sequence, while the response of the later portion of the network decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that the non-target response and $\theta$ contrasts were robustly positively correlated (median Pearson's $\rho$ 0.992, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast (Figure \ref{fig6}c) and the median contrast in $\theta$ at the start of B trials (Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in \dev{} than \msc{} (T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). The contrast in $\theta$ followed the same pattern, with the early portion exhibiting stronger TA in \dev{} (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).

Finally, to show that early $\theta$ was causal in reducing the response to B, we correlated the median contrast in $\theta$ among the early portion of neurons with the network-wide median contrast of the response to stimulation in B, as shown in Figure \ref{fig6}e. Across networks and stimuli, we found a strong negative correlation ($\rho$ = -0.587, p = 9.56e-13, Wald test, n = 120), indicating that greater TA in the neurons responding early was indeed predictive of a lowered response to B in the overall network.

To sum up, we showed in Figure \ref{fig4} that greater $\theta$ in the portion of the network responding earliest to a stimulus plays an important role in reducing the overall response. Here, we showed that the neurons responding earliest to B exhibit higher $\theta$ in the \dev{} sequence as a consequence of an increased response, and that this increase in TA was directly responsible for reducing the response size and TA levels of the network as a whole.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig7.pdf}
    \caption{}
    \label{fig7}
\end{figure*}
\textbf{Fig. 7} Adding STD to the model maintains or slighly enhances the effects of TA.
\textbf{a} Median contrast (\dev{} - \msc{}) in the non-target response (left) and in $\theta$ at the start of B trials (right) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to Figure \ref{fig6}c-d.
\textbf{b} Median contrast (\dev{} - \msc{}) in the non-target response magnitude across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (\dev{} - \msc{}) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
\textbf{c} Non-target response contrast, plotted against its Pearson correlation coefficient with the corresponding contrast in $\theta$ at the start of target trials, in the full model. Compare to Figure \ref{fig5}d.
\textbf{d} Median contrast (\dev{} - \msc{}) in $\theta$ at the start of target trials across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that $\theta$ tends to be reduced more in the full model.
\textbf{e} $\Delta \theta$ at the start of target trials, plotted against its Pearson correlation coefficient with the corresponding contrast in the target response magnitude, in the full model. Compare to Figure \ref{fig4}e.
\textbf{f} Median contrast (\dev{} - \msc{}) in the target response magnitude across networks and stimuli in the full and ablated models, arranged as in panel b. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model

To understand how STD affects deviance detection and causes this difference in response reduction, we first examined the effects of TA in the full model, following the analysis on the ablated model laid out in the previous section. Starting at the cause of the reduced response in non-target trials, we show in Figure \ref{fig7}a the full model equivalent of Figure \ref{fig6}c and d. The contrasts (\dev{} - \msc{}) in the non-target response magnitude and in the resulting $\theta$ levels at the start of B trials both retained a clear separation between the portion of neurons responding early, which showed increased responses (T = 5.44e+03, p = 1.02e-06; one-sided Wilcoxon, n = 120) and increased TA levels (T = 3.41e+03, p = 9.51e-08), and the neurons responding later, whose response (T = 282, p = 9.19e-19) and TA levels (T = 72, p = 2.49e-17) decreased in the \dev{} sequence. Likewise, the correlation across excitatory neurons between these two contrasts remained positive (T = 7.26e+03, p = 9.86e-22) and very strong (median Pearson's $\rho$ = 0.992). This indicates that local suppression to B mediated by TA remained a key driver of the reduced response to non-target trials.

Comparing the non-target response reductions in Figure \ref{fig7}b, we found that they were on average larger in the full model than in the ablated model (T = 4.16e+03, p = 0.0175; one-sided Wilcoxon, n = 120), with the grand medians differing by 36\%. Based on the ablated model analysis in Figure \ref{fig5}, we expected reduced non-target responses to similarly reduce the TA levels at the start of A trials. Indeed, the non-target response contrast remained tightly correlated with the contrast (\dev{} - \msc{}) in $\theta$ at the start of target trials (median Pearson's $\rho$ = 0.989), as shown in Figure \ref{fig7}c. Correspondingly, $\theta$ was reduced in \dev{}, as shown in Figure \ref{fig7}d, and this too by a slightly greater amount in the full model than in the ablated model (T = 4.37e+03, p = 0.0257).

Finally, based on the ablated model analysis in Figure \ref{fig4}, we sought to confirm that lowered $\theta$ before target trials was responsible for the increased target response. As shown in Figure \ref{fig7}e, the contrasts (\dev{} - \msc{}) in $\theta$ at the start of target trials and in response magnitude to A were negatively correlated (T = 30, p = 2.09e-21, median = -0.268; one-sided Wilcoxon, n = 120), though the correlation was weaker in the full model (T = 1.67e+03, p = 1.49e-07). Despite the weaker correlation, however, the full model target response was clearly increased in \dev{} over \msc{}, as shown in Figure \ref{fig7}f, and this increase was greater in the full model than in the ablated model (T = 1.94e+03, p = 0.000723), with the grand medians differing by 94\%. This greater contrast, of course, directly reflects the greater DDI reported in Figure \ref{fig3}.
